Baseline centering problem

### 2018-12-19 paper/tangotd.tex

\paragraph{Better variant?}
I think
everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
that
\begin{equation}
\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} Z(s_1,s_2)
\psi(s_2)
\end{equation}
and this is still linear in $\psi$. This represents the inverse Laplace
operator by
\begin{equation}
Z=\sum_{s_1,s_2}\rho(s_2)  Z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
\end{equation}
In particular the value function is
\begin{equation}
V(s_1)=\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)
\end{equation}
Downside: this representation of the value function does not allow for first
computing $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
$\green$. Here we would have to learn an auxiliary network to imitate
$\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
problem). Alternatively, policy gradient can just directly be obtained
via
\begin{equation}
\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} Z(s',s_2)R(s_2)
\end{equation}
(this works even if $\rho$ is \emph{not} the invariant distribution). We
can apply this each time a new reward is observed at some point $s_2$. We
can even set another arbitrary distribution on $s_1$: this policy
gradient will maximize
the reward obtained from a starting point from that distribution.

The left Bellman equation is
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}
\,Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
\E_{s_2\sim \rho}\,
Z(s',s_2)\psi(s_2)
\end{gather}
for any $s_1$ and $\psi$.
The right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}
Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
Z(s_1,s_2)\psi(s'_2)
\end{gather}
These equations impose a kind of smoothness on $Z$ with respect to the
random walk $P$.

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right yields
\begin{multline}
Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}
Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho, s'_3\sim P(s'_3|s_3)}
Z(s_1,s_3)Z(s'_3,s_2)
\end{multline}
and we should find a way to initialize $Z$ so that $Z\approx \Id$ at time
$0$ ($Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
enough). In the end this Newton equation alone might be sufficient to
encompass the left and right Bellman. It does not rely on test functions,
but the smoothing is implicit via the model used for $Z(s_1,s_2)$.

>>>>

\paragraph{Better variant?}
I think
everything works equally well if we just learn a kernel $Z(s_1,s_2)$ such
that
\begin{equation}
\transp{\1_{s_1}}\Delta^{-1}\psi\approx \E_{s_2\sim \rho} \,Z(s_1,s_2)
\psi(s_2)
\end{equation}
and this is still linear in $\psi$. This represents the inverse Laplace
operator by
\begin{equation}
Z=\sum_{s_1,s_2}\rho(s_2)  Z(s_1,s_2)\1_{s_1}\transp{\1_{s_2}}
\end{equation}
In particular the value function is
\begin{equation}
V(s_1)=\E_{s_2\sim \rho} \, Z(s_1,s_2) R(s_2)
\end{equation}
Downside: this representation of the value function does not allow for first
computing a representation of $R$ as $\E_{s_2} R(s_2)f(s_2)$ online, then taking a dot product with
$\green$. Here we would have to learn an auxiliary network to imitate
$\E_{s_2\sim \rho} Z(s_1,s_2) R(s_2)$ (but this is a supervised learning
problem). Alternatively, policy gradient can just directly be obtained
via
\begin{equation}
\partial \pi(a|s)\E_{s'|s,a} \E_{s_2\sim \rho} \, Z(s',s_2)R(s_2)
\end{equation}
(this works even if $\rho$ is \emph{not} the invariant distribution). We
can apply this each time a new reward is observed at some point $s_2$. We
can even set another arbitrary distribution on $s_1$: this policy
gradient will maximize
the reward obtained from a starting point from that distribution.

The left Bellman equation is
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}
\,Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}
\E_{s_2\sim \rho}\,
Z(s',s_2)\psi(s_2)
\end{gather}
for any $s_1$ and $\psi$.
The right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\E_{s_2\sim \rho}\,
Z(s_1,s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}\,
Z(s_1,s_2)\psi(s'_2)
\end{gather}
These equations impose a kind of smoothness on $Z$ with respect to the
random walk $P$.

The Newton update/middle Bellman equation is
\begin{equation}
Z\gets (1+\eta)Z-\eta Z^2+\eta \gamma ZPZ
\end{equation}
which after applying to $\1_{s_1}$ on the left and $\1_{s_2}/\rho(s_2)$
on the right yields
\begin{multline}
Z(s_1,s_2)\gets (1+\eta)Z(s_1,s_2)-\eta \E_{s_3\sim \rho}\,
Z(s_1,s_3)Z(s_3,s_2)\,+\\\eta \gamma \E_{s_3\sim\rho,\, s'_3\sim
P(s'_3|s_3)}\,
Z(s_1,s_3)Z(s'_3,s_2)
\end{multline}
which is perfectly samplable. We can simulate the assignment
$Z(s_1,s_2)\gets...$ via gradient descent on the parameters of $Z$.

We should find a way to initialize $Z$ so that $Z\approx \Id$ at time
$0$, namely $Z(s,s)=1/\rho(s)$, although starting with $Z(s,s)=1$ should be
quite fine. $Z>0$ (unless we work with $\gamma=1$ and relative stuff
TODO), so the parameterization could reflect that.

In the end this Newton equation alone might be sufficient to
encompass the left and right Bellman. It does not rely on test functions,
but the smoothing is implicit via the model used for $Z(s_1,s_2)$.

Newton has a potential to be numerically less stable: first because of
the $(1+\eta)$ factor which makes $Z$ quickly grow when it's small (until
the $-Z^2$ kicks in), then because of the $-Z^2$ factors which can be
problematic when $Z$ is large. Small enough learning rates should be
stable.
